---
title: "Problem Set 4 Help"
author: "Jenni Putz"
date: "3/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup
```{r}
library(pacman)
p_load(tidyverse, lmtest)

df <- read_csv("/Users/jenniputz/Downloads/004-data.csv")
```

## Testing for Autocorrelation

```{r}
reg <- lm(price_coal ~ lag(price_coal, 1) + lag(price_coal, 2) + lag(price_coal, 3) + 
            price_gas + lag(price_gas, 1) + lag(price_gas, 2) + lag(price_gas, 3), data = df)
summary(reg)
```

Next, get the residuals - use 3 NAs because we have 3rd order lags.

```{r}
df$e <- c(NA, NA, NA, residuals(reg))
```

Next run the Breusch-Godfrey test. We will check for third-order autocorrelation.

```{r}
bg_reg <- lm(e ~ price_coal + lag(price_coal, 1) + lag(price_coal, 2) + lag(price_coal, 3) + 
            price_gas + lag(price_gas, 1) + lag(price_gas, 2) + lag(price_gas, 3) +
            lag(e, 1) + lag(e, 2) + lag(e, 3), data = df)

summary(bg_reg)
```


We dont want the lag(e, n) coefficients to be statistically significant because then there is a relationship between the lagged residuals and the residuals (this is autocorrelation!). Can check it with the `waldtest` function we used in lab.

```{r}
p_load(lmtest)
#waldtest(bg_reg, c("lag(e, 1)", "lag(e, 2)", "lag(e, 3)"))

# NOTE: if the waldtest function doesn't work for you (like it didn't for me), try using this one from the survey package:
p_load(survey)
regTermTest(bg_reg, c("lag(e, 1)", "lag(e, 2)", "lag(e, 3)"), null=NULL,df=NULL, method=c("Wald"))
# you can find the p-value in the bottom right corner of this output
```

Remember: $H_0: r_1 = r_2 = r_3 = 0$ so we reject the null hypotheis and conclude that at least one of the coefficients on the lagged residuals is not equal to zero.